---
title: "Reproducible Data Project 2"
author: "chrisgoose"
date: "8/14/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999) # make knitr print large numbers expanded form
```

# The Effects of Storm & Severe Weather on Economics & Population Health


## Intro

Storms and other severe weather events can cause both public health and economic 
problems for communities and municipalities. Many severe events can result in 
fatalities, injuries, and property damage, and preventing such outcomes to the 
extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric 
Administration's (NOAA) storm database. This database tracks characteristics of 
major storms and weather events in the United States, including when and where 
they occur, as well as estimates of any fatalities, injuries, and property damage.

The basic goal of this assignment is to explore the NOAA Storm Database and 
answer some basic questions about severe weather events.

My data analysis addresses the following questions:

1. Across the United States, which types of events are most harmful with respect 
to population health?

2. Across the United States, which types of events have the greatest economic 
consequences?

This report could be read by a government or municipal manager who might be 
responsible for preparing for severe weather events and will need to prioritize 
resources for different types of events. However, there are no specific 
recommendations in my report.


## Data Processing

First, I downloaded the NOAA storm database and extracted the file. I removed 
most of the variables to create a lower dimension dataset. I kept  variables 
related to event type, number of fatalities, number of injuries, value of 
property damage, and value of crop damage.

```{r get data}
##Here are the data for the project:
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

if(!file.exists("repdata_data_StormData.csv.bz2")) {
        download.file(url, "repdata_data_StormData.csv.bz2", method="curl")
}
stormData <- read.csv("repdata_data_StormData.csv.bz2")

colnames(stormData)

#create lower dimension table
stormData2 <- stormData[, c("REFNUM", "EVTYPE", "FATALITIES", "INJURIES", 
                            "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]

str (stormData2)
```


Next, I handled conversions of the property and crop damage variables, with
support from this [anonmyous RPub doc on the ^EXP labels](https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html). 
In general, labels such as B, M, K, H are converted to 9, 6, 3, 2, respectively,
to indicate an exponential power of 10. From there, the values in the property 
and crop damage columns are raised to the power of 10 in the ^EXP column.

```{r clean data}

#change lowercase labels to uppercase, i.e., 'm' to 'M'
stormData2$PROPDMGEXP <- toupper(stormData2$PROPDMGEXP)
stormData2$CROPDMGEXP <- toupper(stormData2$CROPDMGEXP)

#change alpha & symbolic labels to numeric for exponents
stormData2[stormData2$PROPDMGEXP == "B","PROPDMGEXP"] <- 9
stormData2[stormData2$PROPDMGEXP == "M","PROPDMGEXP"] <- 6
stormData2[stormData2$PROPDMGEXP == "K","PROPDMGEXP"] <- 3
stormData2[stormData2$PROPDMGEXP == "H","PROPDMGEXP"] <- 2
stormData2[stormData2$PROPDMGEXP %in% c("+","-","?","_", ""),"PROPDMGEXP"] <- 0

stormData2[stormData2$CROPDMGEXP == "B","CROPDMGEXP"] <- 9
stormData2[stormData2$CROPDMGEXP == "M","CROPDMGEXP"] <- 6
stormData2[stormData2$CROPDMGEXP == "K","CROPDMGEXP"] <- 3
stormData2[stormData2$CROPDMGEXP == "H","CROPDMGEXP"] <- 2
stormData2[stormData2$CROPDMGEXP %in% c("+","-","?","_", ""),"CROPDMGEXP"] <- 0

unique(c(stormData2$PROPDMGEXP, stormData2$CROPDMGEXP))

stormData2$PROPDMG <- with(stormData2, PROPDMG * 10^as.numeric(PROPDMGEXP))
stormData2$CROPDMG <- with(stormData2, CROPDMG * 10^as.numeric(CROPDMGEXP))

stormData2 <- stormData2[, c("REFNUM", "EVTYPE", "FATALITIES", "INJURIES", 
                            "PROPDMG", "CROPDMG")]
names(stormData2)[1:6] <- c("id", "eventType", "fatalities", "injuries", 
                            "propertyDamage", "cropDamage")
head(stormData2)
```


## Results

### Population Health
To begin exploring the general equation "Across the United States, which types 
of events are most harmful with respect to population health?", I decided to
answer the specific questions:

* Which types of events had the most fatalities?

* Which types of events had the most injuries?

#### Most fatal events.

The top 15 "most fatal" events are ranked below; the top 10 are shown in the 
bar chart. Storm events appear to follow a Pareto distribution in regard to 
injuries. 

```{r fatalities}
#make tidy fatalities data, then rank events by sum
fatality <- stormData2[, c("id", "eventType","fatalities")]
fatalitySum <- aggregate(fatality$fatalities, 
                         by = list(eventType = fatality$eventType), sum)
fatalitySum <- fatalitySum[order(fatalitySum$x, decreasing = TRUE), ]
head(fatalitySum, 15)

#make barplot
require(ggplot2)
ggplot(fatalitySum[1:10, ], aes(x = reorder(eventType, -x), y = x)) + 
        geom_bar(stat = "identity") + xlab("Storm Event Type") + 
        ylab("Sum of Fatalities") + ggtitle("Fatalities by Event, 1950-2011") + 
        theme(axis.text.x = element_text(angle = 90))

#tornado > next 5
fatalitySum$x[[1]]
sum(fatalitySum[2:6,2])
```

Tornados (`r fatalitySum$x[[1]]` injuries) have caused the 
most injuries, more than the next 5 events combined 
(`r sum(fatalitySum[2:6,2])` injuries).


#### Most injurious events

The top 15 "most injurious" events are ranked below; the top 10 are shown in the 
bar chart. Storm events again appear to follow a Pareto distribution, this time
in regard to injuries.

```{r injuries}
#make tidy injuries data, then rank event by sum
injury <- stormData2[, c("id", "eventType","injuries")]
injurySum <- aggregate(injury$injuries, by = list(eventType = injury$eventType), 
                       sum)
injurySum <- injurySum[order(injurySum$x, decreasing = TRUE), ]
head(injurySum, 15)

#make barplot
ggplot(injurySum[1:10, ], aes(x = reorder(eventType, -x), y = x)) + 
        geom_bar(stat = "identity") + xlab("Storm Event Type") + 
        ylab("Sum of Injuries") + ggtitle("Injuries by Event, 1950-2011") + 
        theme(axis.text.x = element_text(angle = 90))

#tornado > all else
injurySum$x[[1]]
sum(injurySum$x) - injurySum$x[[1]]
```

Tornados (`r injurySum$x[[1]]` injuries) have caused the 
most injuries, more than all other events combined 
(`r sum(injurySum$x) - injurySum$x[[1]]` injuries).


### Economic Damage

To begin exploring the general equation "Across the United States, which types 
of events have the greatest economic consequences?", I decided to answer the 
specific questions:

* Which types of events caused the highest costs in crop damage?

* Which types of events caused the highest costs in property damage?

* Which types of events caused the highest combined costs of economic damage?

#### Most economic damage to crops

The top 15 events causing crop damage are ranked below. Storm events again 
appear to follow a Pareto distribution, now with economic loss and crop damage.

```{r crop damage}
#make tidy crop damage data, then rank events by sum
crop <- stormData2[, c("id", "eventType","cropDamage")]
cropSum <- aggregate(crop$cropDamage, by = list(eventType = crop$eventType), 
                       sum)
cropSum <- cropSum[order(cropSum$x, decreasing = TRUE), ]
head(cropSum, 15)

#drought > next 2
cropSum$x[[1]]
sum(cropSum[2:3,2])
```

Drought (`r cropSum$x[[1]]` dollars) have caused the most crop damage, more than 
the next two events (flood & river flood) combined ($`r sum(cropSum[2:3,2])`).


#### Most economic damage to property

The top 15 events causing property damage are ranked below. Yet another instance 
of a Pareto distribution, now with economic loss and property damage.

```{r property damage}
#make tidy property damage data, then rank events by sum
property <- stormData2[, c("id", "eventType","propertyDamage")]
propertySum <- aggregate(property$propertyDamage, 
                         by = list(eventType = property$eventType), sum)
propertySum <- propertySum[order(propertySum$x, decreasing = TRUE), ]
head(propertySum, 15)

#flood > next 2
propertySum$x[[1]]
sum(propertySum[2:3,2])
```

Floods (`r propertySum$x[[1]]` dollars) have caused the most crop damage, more 
than the next two events (hurricane & tornado) combined 
($`r sum(propertySum[2:3,2])`).


#### Most economic damage to crops and property combined

The top 15 events causing combined economic damage are ranked below; the bar plot
shows the top 10 events. Yet another instance of a Pareto distribution, now with 
combined economic loss.

```{r economic damage}
#add prop & crop damage, then rank events by sum
econ <- stormData2[, c("id", "eventType","cropDamage", "propertyDamage")]
econ$econDamage <- with(econ, econ$cropDamage + econ$propertyDamage)
econSum <- aggregate(econ$econDamage, by = list(eventType = econ$eventType), 
                       sum)
econSum <- econSum[order(econSum$x, decreasing = TRUE), ]
head(econSum, 15)

#make barplot
require(ggplot2)
ggplot(econSum[1:10, ], aes(x = reorder(eventType, -x), y = x)) + 
        geom_bar(stat = "identity") + xlab("Storm Event Type") + 
        ylab("Sum of Economic Damage ($)") + 
        ggtitle("Economic Property & Crop Damage by Event, 1950-2011") + 
        theme(axis.text.x = element_text(angle = 90))

#drought > next 2
econSum$x[[1]]
sum(econSum[2:3,2])
```

Floods (`r econSum$x[[1]]` dollars) have caused the most economic damage, 
more than the next two events (hurricane & tornado) combined 
($`r sum(econSum[2:3,2])`).


## Conclusion

* Tornados have caused the most damage to population health if measuring 
fatalities and injuries. Tornados have also imposed the 3rd largest cost in 
economic loss.
* Droughts have imposed the largest cost in economic loss measuring crop damage.
* Floods have imposed the largest cost in economic loss, and when isolating
property damage. Floods have also imposed the second largest crop damage. The 
similar "flash floods" have imposed the sixth largest crop damage. "River 
floods" have imposed the ninth highest property damage.